# Understanding predictors of adult blood pressure using generalized additive modeling (GAM)

Generalized additive models allow us to specify that certain quantitative predictors in a regression model are represented with splines to accommodate non-linearity.  The spline degrees of freedom is determined automatically, and a smoothing penalty is used to further optimize the bias/variance tradeoff.  Further, inference procedures have been devised to properly account for the tuning process.

The "additive" in generalized additive models suggests that these models are additive in the sense of not having interactions.  However, it is also possible to incorporate interactions into a GAM, in which case the models are no longer additive.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(mgcv)
```

```{r}
source("read.R")
```

```{r}
dx = df %>% dplyr::select(BPXSY1, RIAGENDR, RIDAGEYR, BMXWT, BMXHT, BMXBMI, BMXLEG, BMXARML, BMXARMC, BMXWAIST, BMXHIP)
dx = dx %>% drop_na()
```

```{r}
dx$RIAGENDRx = recode(dx$RIAGENDR, "F"=1, "M"=-1)
dx$RIAGENDR = as.factor(dx$RIAGENDR)

dx = dx %>% mutate(BMXBMI=pmax(15, pmin(BMXBMI, 50)))
```

Compare sexes at fixed BMI

```{r}
plot1 = function(mm, bmi, ii) {
    dp = dx[1:100,]
    dp$RIDAGEYR = seq(18, 80, length.out=100)
    dp$BMXBMI = bmi

    dr = data.frame()
    for (sex in c("F", "M")) {
        dp$RIAGENDR = sex
        sbp = predict(mm, newdata=dp)
        d2 = data.frame(age=dp$RIDAGEYR, sbp=sbp, sex=sex)
        dr = rbind(dr, d2)
    }

    plt = ggplot(aes(x=age, y=sbp, color=sex, by=sex), data=dr) + geom_line()
    plt = plt + ggtitle(sprintf("Model %d (BMI=%.0f)", ii, bmi))
    plt = plt + labs(x="Age", y="SBP")
    print(plt)
}
```

Compare BMI 30 versus 25 for one sex.

```{r}
plot2 = function(mm, sex, ii) {
    dp = dx[1:100,]
    dp$RIDAGEYR = seq(18, 80, length.out=100)
    dp$RIAGENDR = sex

    dr = data.frame()
    for (bmi in c(25, 30)) {
        dp$BMXBMI = bmi
        sbp = predict(mm, newdata=dp)
        d2 = data.frame(age=dp$RIDAGEYR, sbp=sbp, bmi=dp$BMXBMI)
        dr = rbind(dr, d2)
    }
    dr$bmi = as.factor(dr$bmi)

    plt = ggplot(aes(x=age, y=sbp, color=bmi, by=bmi), data=dr) + geom_line()
    plt = plt + ggtitle(sprintf("Model %d (%s only)", ii, sex))
    plt = plt + labs(x="Age", y="SBP")
    print(plt)
}
```

```{r}
plot_all = function(mm, ii) {
    plot(mm, se=TRUE)
    plot1(mm, 25, ii)
    plot1(mm, 30, ii)
    plot2(mm, "F", ii)
    plot2(mm, "M", ii)
}
```

A simple additive model

```{r}
f0 = as.formula("BPXSY1 ~ s(RIDAGEYR) + s(BMXBMI) + RIAGENDR")
m0 = gam(f0, data=dx)
plot_all(m0, 0)
```

Allow the age effects to differ by sex

```{r}
f1 = as.formula("BPXSY1 ~ s(RIDAGEYR, by=RIAGENDR) + BMXBMI")
m1 = gam(f1, data=dx)
plot_all(m1, 1)
```

Allow the BMI effects to differ by sex

```{r}
f2 = as.formula("BPXSY1 ~ RIDAGEYR + s(BMXBMI, by=RIAGENDR)")
m2 = gam(f2, data=dx)
plot_all(m2, 2)
```

Allow both the age and BMI effects to differ by sex

```{r}
f3 = as.formula("BPXSY1 ~ s(RIDAGEYR, by=RIAGENDR) + s(BMXBMI, by=RIAGENDR)")
m3 = gam(f3, data=dx)
plot_all(m3, 3)
```

Allow age and BMI to relate non-additively to SBP

```{r}
f4 = as.formula("BPXSY1 ~ te(RIDAGEYR, BMXBMI) + s(RIDAGEYR, by=RIAGENDR) + s(BMXBMI, by=RIAGENDR)")
m4 = gam(f4, data=dx)
plot_all(m4, 4)
```

Allow the non-additive relationship between age and BMI to differ by sex

```{r}
f5 = as.formula("BPXSY1 ~ te(RIDAGEYR, BMXBMI, by=RIAGENDR) + s(RIDAGEYR, by=RIAGENDR, sp=1.5) + s(BMXBMI, by=RIAGENDR)")
m5 = gam(f5, data=dx)
plot_all(m5, 5)
```

# Statistical inference

Add a completely random covariate to see how well the inference works

```{r}
dx$E = rnorm(dim(dx)[1])
f6 = as.formula("BPXSY1 ~ te(RIDAGEYR, BMXBMI, by=RIAGENDR) + s(RIDAGEYR, by=RIAGENDR) + s(BMXBMI, by=RIAGENDR) + s(E)")
m6 = gam(f6, data=dx)
```

Compare each successive pair of models.

```{r}
ml = list(m0, m1, m2, m3, m4, m5, m6)
for (i in 1:(length(ml)-1)) {
    aa = anova(ml[[i]], ml[[i+1]], test="F")
    cat(sprintf("Model %d versus %d: %f\n", i-1, i, aa[["Pr(>F)"]][2]))
}
```

