# Analyzing demographic variation among US counties using biplots

The data considered here are a single year of population counts for US counties. The population within each county is partitioned into 2 x 2 x 4 x 19 = 304 demographic cells (sex тип Hispanic ethnicity status тип race тип age). See the prep.R script for more information.

In this notebnook we will use principal components analysis (PCA) and biplots to understand the covariation of demographic group sizes across US counties.

```{r}
library(ggplot2)
library(stringr)
library(purrr)
library(dplyr)
```

```{r}
source("prep.R")
```

Only consider counties for which we have natality data:

```{r}
fips = sort(unique(births$FIPS))
fips = intersect(fips, demog$FIPS)
demog = demog %>% filter(FIPS %in% fips)
demog = demog %>% arrange(demog, FIPS)
```

Restructure the data into long form:

```{r}
dl = pivot_longer(select(demog, -FIPS), cols=everything(), names_to="group", values_to="population")
dl = separate(dl, group, c("race", "ethnicity", "sex", "age"), sep="_")
ds = group_by(dl, race, ethnicity, sex) %>% summarize(population=sum(population))
ds
```

Create a data matrix omitting the FIPS index column

```{r}
demogx = demog %>% select(-FIPS)
demogx = as.matrix(demogx)
```

The total population per county.

```{r}
totpop = apply(demogx, 1, sum)
```

Log transform the population counts for variance stabilization and symmetry.

```{r}
demogx = log(1 + demogx)
totpop = log(1 + totpop)
```

Double center the data

```{r}
gmn = mean(demogx)
demogx = demogx - gmn
colmn = apply(demogx, 2, mean)
demogx = t(t(demogx) - colmn)
rowmn = apply(demogx, 1, mean)
demogx = demogx - rowmn
```

Factor the matrix

```{r}
ss = svd(demogx)
```

Extract the components of the SVD

```{r}
u = ss$u
s = ss$d
v = ss$v
```

Plot the singular values to assess the dimensionality:

```{r}
dp = data.frame(x=seq(length(s)), y=s)
plt = ggplot(data=dp, aes(x=x, y=y)) + geom_line() + geom_point()
plt = plt + xlab("Position") + ylab("Singular value")
print(plt)
```
Consider exponential and power law models for the singular values, by making scree plots in semi-log space and log space.  One reasonable interpretation of these plots is that there are around 10-12 relatively larger singular values, and the remainder follow an exponentially decaying pattern.

```{r}
ss = s[1:(length(s)-1)]
dp = data.frame(x=seq(length(ss)), y=ss)
plt = ggplot(data=dp, aes(x=x, y=log(y))) + geom_line() + geom_point()
plt = plt + xlab("Position") + ylab("Log singular value")
plt = plt + ggtitle("Assess fit for exponential model ")
print(plt)
```
```{r}
plt = ggplot(data=dp, aes(x=log(x), y=log(y))) + geom_line() + geom_point()
plt = plt + xlab("Log position") + ylab("Log singular value")
plt = plt + ggtitle("Assess fit for powerlaw model ")
print(plt)
```

Construct the row and column scores

```{r}
#alpha = 1 # distance interpretation for rows (counties)
#alpha = 0 # distance interpretation for columns (demographic categories)
alpha = 0.5
uu = u %*% diag(s^alpha)
vv = v %*% diag(s^(1-alpha))
```

```{r}
generate_biplots = function(uu, vv, sx, j1, j2, highlight=NULL) {
  cn = names(demog)
  cx = str_split(cn, "_")
  cx = cx[2:length(cx)]

  ud = data.frame(u1=uu[,j1], u2=uu[,j2], fips=fips)
  vd = data.frame(v1=vv[,j1], v2=vv[,j2], race=cx%>%map(1)%>%list_simplify, eth=cx%>%map(2)%>%as_vector, sex=cx%>%map(3)%>%as_vector)
  vd1 = vd %>% filter(sex==sx)

  # Plot grey points for the counties
  plt = ggplot(data=ud, aes(x=u1, y=u2)) + geom_point(alpha=0.2)

  # Plot the total county population as a passive variable
  ux = cbind(uu[,j1], uu[,j2])
  pt = solve(t(ux) %*% ux, t(ux) %*% totpop)
  pt = 2 * pt / norm(pt)
  dp = data.frame(x=c(0, pt[1]), y=c(0, pt[2]))
  plt = plt + geom_segment(x=0, y=0, xend=pt[1], yend=pt[2], 
                           arrow=arrow(length=unit(0.5, "cm")))

  if (!is.null(highlight)) {
     uf = left_join(highlight, ud, "fips")
     plt = plt + geom_label(data=uf, aes(x=u1, y=u2, label=label))
  }
    
  # Plot colored points for the demographic variables
  plt = plt + geom_path(data=vd1, aes(x=v1, y=v2, group=interaction(race, eth), color=interaction(race, eth)))

  plt = plt + xlab(sprintf("Component %d", j1)) + ylab(sprintf("Component %d", j2))
  plt = plt + ggtitle(ifelse(sx=="F", "Female", "Male"))
  print(plt)
}
```

Select some counties to highlight in the plots (Wayne county MI, Santa Clara CA, Bristol MA, Cook IL, Pennington SD, Los Angeles CA).

```{r}
f = c("26163", "06085", "25005", "17031", "46103", "06037")
b = c("W", "S", "B", "C", "P", "L")
highlight = data.frame(fips=f, label=b)
```

```{r}
generate_biplots(uu, vv, "F", 1, 2, highlight=highlight)
```

```{r}
generate_biplots(uu, vv, "M", 1, 2, highlight=highlight)
```

```{r}
generate_biplots(uu, vv, "F", 3, 4, highlight=highlight)
```

```{r}
generate_biplots(uu, vv, "M", 3, 4, highlight=highlight)
```