# Demographic predictors of birth counts in US counties from 2011-2020

In this notebook we will explore predictors of natality, defined here as birth count per county/year.  The analysis considers a subset of US counties between 2011 and 2020.

```{r}
library(geepack)
library(ggplot2)
```

```{r}
source("prep.R")
```

Calculate the mean and variance within each county to assess the mean/variance relationship.

```{r}
mv = births %>% group_by(FIPS) %>% summarize(births_mean=mean(Births), births_var=var(Births))
mv = mv %>% mutate(log_births_mean=log(births_mean), log_births_var=log(births_var))
```

Plot the variance against the mean on the log/log scale to assess the mean/variance relationship.

```{r}
plt = ggplot(mv, aes(x=log_births_mean, y=log_births_var)) + geom_point()
print(plt)
```

Use a linear model to assess the mean variance relationship within the family of relationships representable as Var[E[Y|X]] = a*E[Y|X]^b.

```{r}
mmv = lm(log_births_var ~ log_births_mean, data=mv)
summary(mmv)
```

Merge the birth data with population, RUCC, and ADI data

```{r}
da = merge(births, pop, on="FIPS", how="left")
da = merge(da, rucc, on="FIPS", how="left")
da = merge(da, adi, by.x="FIPS", by.y="FIPS5", how="left")
```

We will be using population as an offset below, so log transform it here.

```{r}
da = da %>% mutate(logPop = log(Population))
da = da %>% mutate(logADINatRank = log(ADI_NATRANK))
```

To simplify interpretation of interactions, we create centered versions of quantitative predictors below.

```{r}
da = da %>% mutate(yearc = year - mean(year))
da = da %>% mutate(logPopc = logPop - mean(logPop))
da = da %>% mutate(RUCC_2013c = RUCC_2013 - mean(RUCC_2013))
da = da %>% mutate(logADINatRank = logADINatRank - mean(logADINatRank))
da = da %>% mutate(logADINatRankZ = logADINatRank / sd(logADINatRank))
head(da)
```

Basic GLM looking at births in terms of population and urbanicity. This model does not account for correlations between repeated observations on the same county.

```{r}
r0 = glm(Births ~ logPop + logADINatRankZ + RUCC_2013, quasipoisson(), da)
summary(r0)
```

GEE accounts for the correlated data, no need to use quasipoisson here since GEE is already robust to this form of heteroscedasticity.

```{r}
r1 = geeglm(Births ~ logPop + logADINatRankZ + RUCC_2013, data=da, id=FIPS, family=poisson())
summary(r1)
```

GEE with log population as an offset instead of being a covariate.

```{r}
r2 = geeglm(Births ~ RUCC_2013 + logADINatRankZ, data=da, id=FIPS, family=poisson(), offset=logPop)
summary(r2)
```

GEE with log population as an offset instead of being a covariate.

```{r}
r3 = geeglm(Births ~ RUCC_2013 + logADINatRankZ, data=da, id=FIPS, family=Gamma(link="log"), offset=logPop)
summary(r3)
```

GEE with log population as an offset instead of being a covariate.

```{r}
r4 = geeglm(Births ~ RUCC_2013 + logADINatRankZ, data=da, id=FIPS, family=Gamma(link="log"),
            corstr="exchangeable", offset=logPop)
summary(r4)
```

Add year as a main effect

```{r}
r5 = geeglm(Births ~ RUCC_2013 + logADINatRankZ + year, data=da, id=FIPS, family=Gamma(link="log"),
            corstr="exchangeable", offset=logPop)
summary(r5)
```

Add an interaction between year and RUCC

```{r}
r6 = geeglm(Births ~ (RUCC_2013c + logADINatRankZ) * yearc, data=da, id=FIPS, family=Gamma(link="log"),
            corstr="exchangeable", offset=logPop)
summary(r6)
```

Next we prepare to fit a Poisson model using principal components regression (PCR). The principal components (PC's) will be based on demographic characteristics of each county.

Use a log transformation to variance-stabilize the counts.

```{r}
demogx = as.matrix(demog[,2:dim(demog)[2]])
demogx = log(1 + demogx)
```

Double center the data

```{r}
gmn = mean(demogx)
demogx = demogx - gmn
colmn = apply(demogx, 2, mean)
demogx = t(t(demogx) - colmn)
rowmn = apply(demogx, 1, mean)
demogx = demogx - rowmn
```

Get factors from the demographic data

```{r}
va = colnames(demog)[2:dim(demog)[2]]
sv = svd(demogx)
```

Calculate the proportion of variance explained by each factor.

```{r}
pve = sv$d^2
pve = pve / sum(pve)
```

Put the demographic factors into a dataframe

```{r}
demog_f = data.frame(FIPS=demog$FIPS)
for (k in 1:100) {
    demog_f[,sprintf("pc%d", k)] = sv$u[, k]
}
```

Merge the birth data with demographic data

```{r}
da = merge(da, demog_f, on="FIPS", how="left")
```

Include this number of factors in the next few models

```{r}
npc = 10
```

GLM, not appropriate since we have repeated measures on counties

```{r}
fml = paste("pc", seq(npc), sep="")
fml = paste(fml, collapse=" + ")
fml = sprintf("Births ~ (logPopc + RUCC_2013c + logADINatRankZ) * yearc + %s", fml)
fml = as.formula(fml)
r7 = glm(fml, quasipoisson(), da)
summary(r7)
```

GEE accounts for the correlated data, use logPop as offset

```{r}
r8 = geeglm(fml, data=da, id=FIPS, family=Gamma(link="log"))
summary(r8)
```

Get the fitted values on the log scale, without including the offset. These values can be used to compare two counties as if they had the same population.  Note that the results depend on the number of PCs (npc).

```{r}
adjfit = model.matrix(r5) %*% coef(r5)
```

This function fits a Poisson GLM to the data using 'npc' principal components as explanatory variables.

```{r}
fitmodel = function(npc, use_gee=TRUE) {

    # Construct a model formula
    fml = paste("pc", seq(npc), sep="")
    fml = paste(fml, collapse=" + ")
    fml = sprintf("Births ~ RUCC_2013c * yearc + %s", fml)
    fml = as.formula(fml)

    # A GEE using log population as an offset.
    if (use_gee) {
        r = geeglm(fml, data=da, id=FIPS, family=Gamma(link="log"), offset=da$logPop)
    } else {
        r = glm(fml, data=da, family=Gamma(link="log"), offset=da$logPop)
    }

    # Convert the coefficients back to the original coordinates
    cf = coef(r)
    ii = grep("pc", names(cf))
    cf = cf[ii]
    cf = sv$v[,1:npc] %*% (cf / sv$d[1:npc])
    cf = data.frame(coef=cf)
    cf$Race = ""
    cf$Origin = ""
    cf$Sex = ""
    cf$Age = ""

    # Create a long-form dataframe containing the coefficients and
    # information about what each coefficient refers to.
    na = names(demog)
    na = na[2:length(na)]
    for (i in 1:length(na)) {
        x = strsplit(na[i], "_")[[1]]
        cf[i, "Race"] = x[1]
        cf[i, "Origin"] = x[2]
        cf[i, "Sex"] = x[3]
        cf[i, "Age"] = x[4]
    }
    return(list(coef=cf, model=r))
}
```

Fit models with these various numbers of PCs and plot the coefficients.  Use GLM to get the coefficient estimates for plotting since GEEPack seems to hang on big models with Gamma families.


```{r}
fit_models = function(pcs, use_gee, do_plot) {
    models = list()
    for (npc in pcs) {
        print(npc)
        r = fitmodel(npc, use_gee)
        models[[length(models)+1]] = r$model
        cf = r$coef

        if (do_plot) {
          plt = ggplot(cf, aes(x=Age, y=coef, group=interaction(Race, Origin, Sex), color=Race, lty=Sex, shape=Origin))
          plt = plt + geom_line() + geom_point() + ggtitle(sprintf("%d factors", npc))
          plt = plt + ggtitle(sprintf("%d PCs, cumulative PVE=%.4f", npc, sum(pve[1:npc])))
          print(plt)
        }
    }
    return(models)
}

pcs = c(5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100)
models = fit_models(pcs, FALSE, TRUE)
```

These score tests should use glmtoolbox but currently that doesn't work.

```{r}
pcs = c(5, 10, 20, 30)
models = fit_models(pcs, TRUE, FALSE)

for (j in 2:(length(models)-1)) {
    st = anova(models[[j-1]], models[[j]], test="score")
    p = st[[3]]
    cat(sprintf("%d vs. %d: p=%f\n", pcs[j+1], pcs[j], p))
}
```
