# Understanding variation in lifespans of notable people using the BHHT data

The BHHT (Brief History of Human Time) project provides a dataset about "notable people" based mainly on wikipedia biography articles.

The analyses below focus on lifespans of the people in the BHHT dataset, aiming to understand how lifespans vary based on factors including era of birth, the geographic region where the person lived, and the person's sex and occupation.

This analysis uses survival analysis methods, allowing us to appropriately use information from still-living people.

```{r}
library(dplyr)
library(survival)
library(readr)
library(splines)
library(ggplot2)
library(tidyr)
```

Change the path below as needed to point to the directory holding the data file.

```{r}
pa = "/home/kshedden/mynfs/data/Teaching/bhht"
```

Load the dataset.  Use the latin-1 encoding since there is some non-UTF data in the file.  Add "n_max=####" when developing to reduce the run time, but use as much data as possible to get final results.

```{r}
df = read_csv(file.path(pa, "cross-verified-database.csv.gz"),
              locale=readr::locale(encoding="latin1"), n_max=200000)
```

Create a lifespan variable (years of life).  It will be missing for people who are currently living.

```{r}
df = df %>% mutate(lifespan = death - birth)
print(table(is.na(df$lifespan)))
```

We exclude people born before 1500, there is too little data to gain a meaningful understanding of the distribution of lifespan prior to this year.

```{r}
dx = df %>% filter(birth >= 1500)
```

Retain only the variables needed below.

```{r}
dx = dx %>% select(birth, lifespan, gender, un_region, level1_main_occ)
head(dx)
```

There are a small number of people with missing or "Other" gender but it is too small of a sample from which to draw conclusions.

```{r}
print(table(dx$gender))
dx = dx %>% filter(gender %in% c("Female", "Male"))
```

Drop uninformative occupation codes.

```{r}
dx = dx %>% filter(!(level1_main_occ %in% c("Missing", "Other")))
print(table(dx$level1_main_occ))
```

Note the distribution of geographic regions:

```{r}
print(table(dx$un_region))
```

Censor lifespans at the calendar year 2022.

```{r}
censor_year = 2022
dx = dx %>% mutate(clifespan = ifelse(is.na(lifespan), censor_year - birth, lifespan))
dx = dx %>% mutate(died = is.finite(lifespan))
```

Now we can drop all rows with missing data

```{r}
dx = dx %>% select(-lifespan) %>% drop_na()
```

Create a categorical variable indicating the century in which a person was born.

```{r}
dx = dx %>% mutate(era = floor((birth - 1500) / 100))
```

Note the fairly strong relationship between era of birth and region:

```{r}
table(dx$un_region, dx$era)
```

# Estimates of marginal survival functions

Plot the marginal survival functions for people born in each century. These survival functions are estimated using the product limit (Kaplan-Meier) method.

```{r}
sf = survfit(Surv(clifespan, died) ~ era, dx)
plot(sf, xlab="Age", ylab="Proportion alive", col=1:6, xlim=c(0, 100))
legend(x="topright", c("1500s", "1600s", "1700s", "1800s", "1900s", "2000s"),
       col=1:6, lty=1)
```

Plot the marginal survival functions for females and males:

```{r}
sf = survfit(Surv(clifespan, died) ~ gender, dx)
plot(sf, xlab="Age", ylab="Proportion alive", col=1:2, xlim=c(0, 100))
legend(x="topright", c("Female", "Male"), col=1:2, lty=1)
```

Plot the marginal survival functions stratified by region:

```{r}
sf = survfit(Surv(clifespan, died) ~ un_region, dx)
plot(sf, xlab="Age", ylab="Proportion alive", col=1:5, xlim=c(0, 100))
legend(x="topright", c("Africa", "America", "Asia", "Europe", "Oceania"), col=1:6, lty=1)
```
The results above may be heavily influenced by confounding.  The people from Oceania tended to live more recently, and people who lived more recently tend to have longer lifespans.  Conversely, many of the notable people from the 1500's-1700's are from Europe, and lifespans tended to be shorter in these historical eras. Looking at one factor at a time, it is not clear whether the "driver" of lifespan variation is geography (where a person lived) or time (when a person lived).

# Survival regression analyses

Create a time variable that has year 1500 as the origin.  Use this as a predictor in regression models below so that year 1500 becomes the reference year (i.e. all effects are relative to 1500).

```{r}
dx = dx %>% mutate(time1500 = birth - 1500)
```

Fit a proportional hazards regression model with main effects for year of birth, gender, occupation, and region.

```{r}
fml = "Surv(clifespan, died) ~ time1500 + gender + level1_main_occ + un_region"
m0 = coxph(as.formula(fml), dx)
m0
```

The function below plots a partial regression plot, showing the log hazard ratio (lhr) for mortatlity versus year of birth, controlling for fixed values of other covariates.

```{r}
plot_birthyear_partial = function(dx, rr, gender="Female", level1_main_occ="Leadership", un_region="Asia") {
  dp = dx[1:100,]
  dp = dp %>% mutate(gender = !!gender, level1_main_occ = !!level1_main_occ, un_region = !!un_region)
  b = seq(1500, 2000, length.out=100)
  dp = dp %>% mutate(time = b)
  dp = dp %>% mutate(time1500 = b - 1500)
  dp$lhr = predict(rr, dp)
  print(head(dp))
    
  plt = ggplot(dp, aes(x=time, y=lhr)) + geom_line()
  print(plt)
}
```

The plot below shows how the log hazard ratio for mortality decreases by approximately 2.5 over 500 years.  This is approximately a factor of 12.

```{r}
plot_birthyear_partial(dx, m0)
```

To assess for curvature, consider a quadratic model for year of birth:

```{r}
fml = "Surv(clifespan, died) ~ poly(time1500, 2) + gender + level1_main_occ + un_region"
m1 = coxph(as.formula(fml), dx)
m1
```

```{r}
plot_birthyear_partial(dx, m1)
```

Now consider a cubic model for year of birth.

```{r}
fml = "Surv(clifespan, died) ~ poly(time1500, 3) + gender + level1_main_occ + un_region"
m2 = coxph(as.formula(fml), dx)
m2
```

```{r}
plot_birthyear_partial(dx, m2)
```

Based on AIC there is strong evidence for curvature in the year of birth effect:

```{r}
aic = c(AIC(m0), AIC(m1), AIC(m2))
aic = aic - min(aic)
aic
```

High degree polynomials make poor basis functions so next we use polynomial splines, which are piecewise polynomials.

```{r}
fml = "Surv(clifespan, died) ~ bs(time1500, 4) + gender + level1_main_occ + un_region"
m4 = coxph(as.formula(fml), dx)
m4
```

```{r}
plot_birthyear_partial(dx, m4)
```

```{r}
plot_birthyear_partial_sexdiff = function(dx, rr) {
  dp = dx[1:200,]
  g = c(rep("Female", 100), rep("Male", 100))
  f = c(rep(1, 100), rep(0, 100))
  dp = dp %>% mutate(gender = g, female=f, level1_main_occ = "Leadership", un_region = "Asia")
  b = seq(1500, 2000, length.out=100)
  dp = dp %>% mutate(time = c(b, b))
  dp = dp %>% mutate(time1500 = time - 1500)
  dp$lhr = predict(rr, dp)
  print(head(dp))
    
  plt = ggplot(dp, aes(x=time, y=lhr, by=gender, color=gender)) + geom_line()
  print(plt)
}
```

```{r}
dx = dx %>% mutate(time1500=birth-1500)
fml = "Surv(clifespan, died) ~ bs(time1500, 4)*gender + level1_main_occ + un_region"
m5 = coxph(as.formula(fml), dx)
summary(m5)
```

```{r}
plot_birthyear_partial_sexdiff(dx, m5)
```
Above we considered an interaction between a categorical variable (gender) and a quantitative variable (time) which was represented with splines.  This leads to pairwise interactions between gender and each spline basis function being included in the model.  An alternative approach is to model the main effect of time with a spline, but use a simpler linear specification for the interaction.  This approach is demonstrated below.

```{r}
dx = dx %>% mutate(time1500 = birth - 1500)
dx = dx %>% mutate(female = ifelse(gender == "Female", 1, 0))
fml = "Surv(clifespan, died) ~ bs(time1500, 4) + time1500:female + female + level1_main_occ + un_region"
m6 = coxph(as.formula(fml), dx)
summary(m6)
```

```{r}
plot_birthyear_partial_sexdiff(dx, m6)
```

# Stratification

Fit a sex-stratified proportional hazards regression model, which gives each sex its own baseline hazard function.

```{r}
dx = dx %>% mutate(time1500=birth-1500)
fml = "Surv(clifespan, died) ~ bs(time1500, 4) + level1_main_occ + un_region + strata(gender)"
m7 = coxph(as.formula(fml), dx)
summary(m7)
```

## Baseline hazard functions

Estimate and plot the baseline cumulative hazard function.  Cumulative hazard functions are easy to estimate but somewhat difficult to interpret.  Note that the basehaz function in R returns the cumulative hazard but the column is labeled 'hazard'.

```{r}
bh = basehaz(m0, center=FALSE)
bh = bh %>% filter(time < 100)
plt = ggplot(bh, aes(x=time, y=hazard)) + geom_line() + xlim(0, 100) +
             xlab("Age") + ylab("Cumulative hazard")
print(plt)
```

Plot the estimated baseline cumulative hazard function on the log scale.

```{r}
plt = ggplot(bh, aes(x=time, y=log(hazard))) + geom_line() + xlim(0, 100) +
             xlab("Age") + ylab("Log cumulative hazard")
print(plt)
```

Estimate the baseline hazard function using numerical differentiation and plot it.

```{r}
haz = diff(bh$hazard) / diff(bh$time)
haz = data.frame(hazard=log(haz), time=bh$time[1:(length(bh$time)-1)])
plt = ggplot(haz, aes(x=time, y=hazard)) + geom_smooth(se=F) + xlim(0, 100) +
             xlab("Age") + ylab("log Hazard")
print(plt)
```

Plot the baseline hazard function for each sex.  

```{r}
bh = basehaz(m7, center=F)
bhx = bh %>% group_by(strata) %>% transform(hh=c(diff(hazard) / diff(time), NA))
plt = ggplot(bhx, aes(x=time, y=log(hh), color=strata)) + geom_smooth(se=F) + xlim(0, 100) + xlab("Age") + ylab("Log hazard")
print(plt)
```
