# Understanding the relationships among characteristics of notable people

"Notability" is a social construct with no fixed definition.  The BHHT data contain four main attributes describing the notable people -- sex, occupation, birth era, region.  These characteristics can be used to understand the social construction of notability in different contexts.  For example, we can consider how the joint distribution of sex and occupation varies geographically (by region) and over time.

The characteristics of notable people in the BHHT data are nominal (except birth year, which can be recoded if desired to an ordinal "birth era" variable).  We can thus represent the data as a 4-way contingency table (sex x occupation x birth era x region). This contingency table reflects the joint distribution of the four characteristics in the population of interest.

```{r}
library(dplyr)
library(FactoMineR)
library(tidyr)
library(readr)
```

Change this path as needed to point to the directory holding the data file.

```{r}
pa = "/home/kshedden/mynfs/data/Teaching/bhht"
```

Next we read the data.  The overall dataset has around 2.4 million rows.  During exploration you might want to limit the analysis to the first million rows, then re-run to get final results with the full data.

```{r}
da = read_csv(file.path(pa, "cross-verified-database.csv.gz"), n_max=1000000)
da = da %>% rename(occ=level1_main_occ, reg=un_region, sex=gender)
```

Retain only the variables used in the analysis.

```{r}
dx = da %>% select(birth, occ, sex, reg)
dx = dx %>% drop_na()
```

Few people in the dataset lived before 1500, so for simplicity we exclude them here.

```{r}
dx = dx %>% filter(birth >= 1500)
```

Create a "century of birth" variable

```{r}
dx = dx %>% mutate(bcen = round(birth, -2))
dx = dx %>% select(bcen, occ, sex, reg)
dx = dx %>% mutate(bcen = as.factor(bcen))
```

Remove small groups that are difficult to interpret due to low precision or high confounding.

```{r}
dx = dx %>% filter(occ != "Other" & occ != "Missing" & sex != "Other")
```

We will aim to understand notability by studying the contingency table below:

```{r}
dz = dx %>% group_by(bcen, occ, sex, reg) %>% summarize(n=n())
```

We can calculate Pearson residuals for the contingency table.

```{r}
nx = sum(dz$n)
vx = c("bcen", "occ", "sex", "reg")
for (v in vx) {
  vp = sprintf("%s_p", v)
  dz = dz %>% group_by(.data[[v]]) %>% mutate("{v}_p" := sum(n)/nx)
}
dz = dz %>% mutate(expected=nx*bcen_p*occ_p*sex_p*reg_p)
dz = dz %>% mutate(pearson_resid=(n-expected)/sqrt(expected))
dz = dz %>% arrange(dz, pearson_resid)
```

## Multiple Correspondence Analysis

Multiple Correspondence Analysis is a type of factor analysis for categorical data.  A common use of MCA is to produce biplots that can be used to visualize the joint distribution of several categorical variables.  Here we use MCA to understand the relationships among nominal (categorical) attributes of notable people.  These data can be thought of as a 4-way contingency table (sex x birth era x occupation x region).  The goal of MCA is to understand the structure of this contingency table as a graph.

```{r}
mm = MCA(dx, ncp=4, graph=FALSE)
```

Plot the first two components of the fitted MCA:

```{r}
plt = plot(mm, axes=c(1, 2), invisible="ind")
print(plt)
```

Plot components 3-4 of the fitted MCA:

```{r}
plt = plot(mm, axes=c(3, 4), invisible="ind")
print(plt)
```
In an MCA plot, information about the variables is encoded both in the angles between the variable scores, and in the magnitudes of the variable scores.  A variable that is independent of the other variables will not contribute to the MCA and will have very short magnitude.  We can see this in the example below, where we simulate a variable that is independent of the others and include it in the MCA.

```{r}
f = as.factor(sample(c("1", "2"), dim(dx)[1], replace=T))
dx = dx %>% mutate(fake=f)
mm1 = MCA(dx, ncp=4, graph=FALSE)
plt = plot(mm1, axes=c(1, 2), invisible="ind")
print(plt)
```
