# Inaturalist

In this notebook we use multilevel regression to assess species-specific linear temporal trends in the latitudes of occurrences of species within a class of plants.  In a multilevel regression model, each species has its own slope for the rate of change of latitude over time.  However instead of estimating these slopes as parameters, they are treated as random effects (random slopes), and we estimate the variance of the probability distribution of random slopes.  This variance parameter reflects the dispersion among species of the rates of change, but does not directly tell us anything about the slope (rate of change) for each individual species.

```{r}
library(readr)
library(lme4)
library(dplyr)
library(splines)
library(tidyr)
library(ggplot2)
library(ggfortify)
library(mapdata)
library(RLRsim)
```

Select a class of plants

```{r}
#pclass = "Pinopsida"
pclass = "Polypodiopsida"

pa = sprintf("Plantae_%s.csv.gz", pclass)
pa = file.path("/home/kshedden/data/Teaching/inaturalist", pa)
```

Below we load the data and retain only occurrences after January 1, 2015.

```{r}
da = read_csv(pa)
da = da %>% mutate(eventDate = as.Date(eventDate))
da = da %>% filter(eventDate >= as.Date('2015-01-01'))
da = da %>% select(!elevation) %>% drop_na()
```

The 'day' variable counts days since January 1, 2015.

```{r}
da = da %>% mutate(date1 = as.Date('2015-01-01'))
da = da %>% mutate(day=difftime(eventDate, date1, units="days")/1000)
```

Get the mean latitude for each species

```{r}
dm = da %>% group_by(species) %>% summarize(species_latitude=mean(decimalLatitude))
da = inner_join(da, dm, by="species")
```

Convert longitude to radians

```{r}
da = da %>% mutate(lonrad=pi*decimalLongitude/180)
```

Plot the locations of the occurrences.

```{r}
world = map_data("world")
plt = ggplot() + ggtitle(pclass)
plt = plt + geom_map(data=world, map=world, aes(long, lat, map_id=region, fill="grey"))
plt = plt + geom_point(data=da, aes(decimalLongitude, decimalLatitude), alpha=0.2)
plt = plt + theme(legend.position="none")
print(plt)
```

Fit a multilevel linear regression with random species intercepts. 

```{r}
m1 = lmer(decimalLatitude ~ (1 | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m1)
```
We can extract the estimated variance parameters (standard deviation of the random species intercept and
residual standard deviation) as follows:

```{r}
vc = VarCorr(m1)
print(vc)
```
We can use the multilevel model to estimate the intraclass correlation coefficient (ICC) for the species latitudes.

```{r}
sv = vc$species[1] # the species variance
rv = summary(m1)$sigma^2 # the residual variance
icc = sv / (sv + rv)
print(icc)
```
 
Below we fit a multilevel linear regression with random species intercepts and random day slopes for species.
These two random effects are correlated with each other.  The estimated variance of the random day slope
tells us to what extent different species ranges are moving (in terms of latitude) over time at different
rates.

```{r}
m2 = lmer(decimalLatitude ~ (1 + day | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m2)
```

A key parameter in the output above is the standard deviation of the random day slopes.  If this variance parameter is zero,
the conclusion should be that all species have the same slope (which is given in the mean structure by the fixed slope
for day).  Testing that a variance parameter is zero is a non-regular hypothesis test (since the null hypothesis is on
the boundary of the parameter space).  The likelihood ratio test statistic in this case does not follow a chi-square distribution
under the null hypothesis.  However, inference based on the chi square distribution in this setting will generally
be conservative.  When there is strong evidence against the null hypothesis, we can still be confident in rejecting the
null hypothesis.  Similar guidance applies to the AIC.

```{r}
print(logLik(m1))
print(logLik(m2))
print(logLik(m1) - logLik(m2))
print(AIC(m1)-AIC(m2))
```

Below we plot the joint density of the random slope and random intercept.  The joint density is Gaussian, so we sample a large number of observations from this Gaussian distribution and plot its empirical distribution (based on 2d nonparametric density estimation).

```{r}
vc = VarCorr(m2)
sv = vc$species[1] # the species variance
va = as.matrix(vc$species)
mn = c(0, fixef(m2)[2])
vr = chol(va)
stopifnot(max(abs(va - t(vr) %*% vr)) < 1e-12)
n = 10000
z = array(rnorm(n*2), c(n, 2)) %*% vr + outer(array(1, n), mn)
z = data.frame(intercept=z[,1], slope=z[,2])
plt = ggplot(z, aes(x=intercept, y=slope)) + stat_density_2d()
print(plt)
```

Later we will conduct a formal test that the day (by species) variance is significantly positive. This requires us to fit two alternative models.

```{r}
m3 = lmer(decimalLatitude ~ (0 + day | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m3)
```

```{r}
m4 = lmer(decimalLatitude ~ (0 + day | species) + (1 | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
```

A formal test that the variance parameter is equal to zero is below.  It may be very slow to run so comment it out if these results are not needed.

```{r}
#ss = exactRLRT(m3, m4, m1)
```

Estimate the marginal mean latitude at Detroit's longitude

```{r}
dp = head(da, 100)
dp = dp %>% mutate(day=seq(min(da$day), max(da$day), length.out=100))
dp = dp %>% mutate(lonrad=-pi*83/180) # Detroit longitude
dp$yy = predict(m1, dp, re.form=NA)
```

Plot the marginal mean latitude for plants observed at Detroit's longitude 

```{r}
plt = ggplot(dp, aes(x=day, y=yy)) + geom_line() + xlab("Time (day x 1000)") + ylab("Marginal mean latitude")
print(plt)
```

The next set of cells produces predicted "trend lines" of latitude versus time for each species using the fitted multilevel model.  These trend lines are "partially pooled", so that for species with large sample sizes, the trend line mainly follows the observed data for that species.  But for species with small sample sizes, the intercept and slope for the species are shrunken toward the population mean intercept and slope.

```{r}
rr = ranef(m2)$species
rr$species = row.names(rr)
nn = group_by(da, species) %>% summarise(n=n())
rr = left_join(rr, nn)
rr = rename(rr, "Intercept"="(Intercept)")

plt = ggplot(rr, aes(x=log(n), y=Intercept)) + geom_point()
print(plt)

plt = ggplot(rr, aes(x=log(n), y=day)) + geom_point()
print(plt)
```


First we get a predicted day 0 latitude value for each species.  Technically this is the marginal mean of the predicted day 0 latitudes, marginalizing over the longitude values (which do vary within a species but not by a lot for most species).

```{r}
dx = da
dx$day = 0
dx$Latitude0 = predict(m2, dx, re.form=NA)
dx = dx %>% group_by(species) %>% summarize(Latitude0=mean(Latitude0))
```

Plot species-level linear trends over this time range

```{r}
day1 = as.numeric(min(da$day))
day2 = as.numeric(max(da$day))
```

Extract the species random effects (random day slopes and intercepts)

```{r}
r0 = ranef(m2)$species
r0 = as.data.frame(r0)
r0$species = row.names(r0)
row.names(r0) = NULL
r0 = r0 %>% left_join(group_by(da, species) %>% summarize(species_latitude=mean(species_latitude)))
```

Use the predicted random effects (BLUPs) to predict species-level mean latitudes on the first and last day covered by the dataset.

```{r}
rr = left_join(r0, dx, by="species")
rr = rr %>% mutate(Latitude1=Latitude0+day*(day2-day1))
rr = rr %>% select(species, Latitude0, Latitude1)
```

Reshape the predictions for plotting

```{r}
pr = pivot_longer(rr, cols=Latitude0:Latitude1)
pr = rename(pr, day=name, latitude=value)
pr = pr %>% mutate(day=recode(day, Latitude0='Day0', Latitude1='Day1'))
pr = pr %>% mutate(name=ifelse(day=="Day0", day1, day2))
```

Plot species-level trends in latitude

```{r}
plt = ggplot(pr, aes(x=day, y=latitude, group=species)) + geom_line(alpha=0.2)
```

Plot the population averaged trend line in red.

```{r}
ang = -pi*83/180
f = fixef(m2)
a = f[["(Intercept)"]] + f[["sin(lonrad)"]]*sin(ang) + f[["cos(lonrad)"]]*cos(ang) +
    f["sin(lonrad/2)"]*sin(ang/2) + f["cos(lonrad/2)"]*cos(ang/2)
b = f[["day"]]
plt = plt + geom_abline(aes(intercept=a, slope=b, color="red"))
```

```{r}
plt = plt + xlab("Time (day x 1000)") + ylab("Species latitude")
print(plt)
```
