# Inaturalist

In this notebook we use multilevel regression to assess species-specific temporal trends in the latitudes of occurrences of species within a class of plants.  Each species has its own slope for the conditional mean of latitude given day.  Instead of estimating these slopes as parameters, they are treated as random effects, and we estimate the variance of the distribution of random slopes.

```{r}
library(readr)
library(lme4)
library(dplyr)
library(splines)
library(tidyr)
library(ggplot2)
library(ggfortify)
library(mapdata)
library(RLRsim)
```

Select a class of plants

```{r}
pclass = "Pinopsida"
#pclass = "Polypodiopsida"

pa = sprintf("Plantae_%s.csv.gz", pclass)
pa = file.path("/home/kshedden/data/Teaching/inaturalist", pa)
```

Below we load the data and retain only occurrences after January 1, 2015.

```{r}
da = read_csv(pa)
da = da %>% mutate(eventDate = as.Date(eventDate))
da = da %>% filter(eventDate >= as.Date('2015-01-01'))
da = da %>% select(!elevation) %>% drop_na()
```

The 'day' variable counts days since January 1, 2015.

```{r}
da = da %>% mutate(date1 = as.Date('2015-01-01'))
da = da %>% mutate(day=difftime(eventDate, date1, units="days")/1000)
```

Get the mean latitude for each species

```{r}
dm = da %>% group_by(species) %>% summarize(species_latitude=mean(decimalLatitude))
da = inner_join(da, dm, by="species")
```

Convert longitude to radians

```{r}
da = da %>% mutate(lonrad=pi*decimalLongitude/180)
```

Plot the locations of the occurrences.

```{r}
world = map_data("world")
plt = ggplot() + ggtitle(pclass)
plt = plt + geom_map(data=world, map=world, aes(long, lat, map_id=region, fill="grey"))
plt = plt + geom_point(data=da, aes(decimalLongitude, decimalLatitude), alpha=0.2)
plt = plt + theme(legend.position="none")
print(plt)
```

Fit a multilevel linear regression with random species intercepts. 

```{r}
m1 = lmer(decimalLatitude ~ (1 | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m1)
```
A key parameter in the output above is the standard deviation of the random day slopes, which we can extract as follows

```{r}
v = VarCorr(m1)
print(v)
```
We can use the multilevel model to estimate the intraclass correlation coefficient (ICC) for the species latitudes.

```{r}
sv = v$species[1] # the species variance
rv = summary(m1)$sigma^2 # the residual variance
icc = sv / (sv + rv)
print(icc)
```
 
Below we fit a multilevel linear regression with random species intercepts and random day slopes for species.

```{r}
m2 = lmer(decimalLatitude ~ (1 + day | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m2)
```
```{r}
v = VarCorr(m2)
sv = v$species[1] # the species variance
v = as.matrix(v$species)
mn = c(0, fixef(m2)[2])
r = chol(v)
stopifnot(max(abs(v - t(r) %*% r)) < 1e-12)
n = 10000
z = array(rnorm(n*2), c(n, 2)) %*% r + outer(array(1, n), mn)
z = data.frame(intercept=z[,1], slope=z[,2])
plt = ggplot(z, aes(x=intercept, y=slope)) + stat_density_2d()
print(plt)
```

Later we will conduct a formal test that the day (within species) variance is significantly positive. This requires us to fit two alternative models.

```{r}
m3 = lmer(decimalLatitude ~ (0 + day | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m3)
```

```{r}
m4 = lmer(decimalLatitude ~ (0 + day | species) + (1 | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
```

The formal test is below, it may be very slow to run so comment it out if these results are not needed.

```{r}
#ss = exactRLRT(m3, m4, m1)
```

Estimate the marginal mean latitude at Detroit

```{r}
dp = head(da, 100)
dp = dp %>% mutate(day=seq(min(da$day), max(da$day), length.out=100))
dp = dp %>% mutate(lonrad=-pi*83/180) # Detroit longitude
dp$yy = predict(m1, dp, re.form=NA)
```

Plot the marginal mean latitude

```{r}
plt = ggplot(dp, aes(x=day, y=yy)) + geom_line() + xlab("Day (x1000)") + ylab("Marginal mean latitude")
print(plt)
```

Get a day 0 value for each species, based on its longitude

```{r}
dx = da
dx$day = 0
dx$Latitude0 = predict(m2, dx, re.form=NA)
dx = dx %>% group_by(species) %>% summarize(Latitude0=mean(Latitude0))
```

Plot species-level linear trends over this time range

```{r}
day1 = as.numeric(min(da$day))
day2 = as.numeric(max(da$day))
```

Species random effects (random day slopes and intercepts)

```{r}
r0 = ranef(m2)$species
r0 = as.data.frame(r0)
r0$species = row.names(r0)
row.names(r0) = NULL
r0 = r0 %>% left_join(group_by(da, species) %>% summarize(species_latitude=mean(species_latitude)))
```

Use the predicted random effects (BLUPs) to predict species-level mean latitudes on the first and last day covered by the dataset.

```{r}
rr = left_join(r0, dx, by="species")
rr = rr %>% mutate(Latitude1=Latitude0+day*(day2-day1))
rr = rr %>% select(species, Latitude0, Latitude1)
```

Reshape the predictions for plotting

```{r}
pr = pivot_longer(rr, cols=Latitude0:Latitude1)
pr = rename(pr, day=name, latitude=value)
pr = pr %>% mutate(day=recode(day, Latitude0='Day0', Latitude1='Day1'))
pr = pr %>% mutate(name=ifelse(day=="Day0", day1, day2))
```

Plot species-level trends in latitude

```{r}
plt = ggplot(pr, aes(x=day, y=latitude, group=species)) + geom_line(alpha=0.5)
```

Plot the population averaged trend line in red.

```{r}
ang = -pi*83/180
f = fixef(m2)
a = f[["(Intercept)"]] + f[["sin(lonrad)"]]*sin(ang) + f[["cos(lonrad)"]]*cos(ang) +
    f["sin(lonrad/2)"]*sin(ang/2) + f["cos(lonrad/2)"]*cos(ang/2)
b = f[["day"]]
plt = plt + geom_abline(aes(intercept=a, slope=b, color="red"))
```

```{r}
plt = plt + xlab("Day (x1000)") + ylab("Species latitude")
print(plt)
```
