# Inaturalist

In this notebook we use multilevel regression to assess species-specific linear temporal trends in the latitudes of occurrences of species within a class of plants.  In a multilevel regression model, each species has its own slope for the conditional mean of latitude given day.  However instead of estimating these slopes as parameters, they are treated as random effects (random slopes), and we estimate the variance of the distribution of random slopes.

```{r}
library(readr)
library(lme4)
library(dplyr)
library(splines)
library(tidyr)
library(ggplot2)
library(ggfortify)
library(mapdata)
library(RLRsim)
```

Select a class of plants

```{r}
#pclass = "Pinopsida"
pclass = "Polypodiopsida"

pa = sprintf("Plantae_%s.csv.gz", pclass)
pa = file.path("/home/kshedden/data/Teaching/inaturalist", pa)
```

Below we load the data and retain only occurrences after January 1, 2015.

```{r}
da = read_csv(pa)
da = da %>% mutate(eventDate = as.Date(eventDate))
da = da %>% filter(eventDate >= as.Date('2015-01-01'))
da = da %>% select(!elevation) %>% drop_na()
```

The 'day' variable counts days since January 1, 2015.

```{r}
da = da %>% mutate(date1 = as.Date('2015-01-01'))
da = da %>% mutate(day=difftime(eventDate, date1, units="days")/1000)
```

Get the mean latitude for each species

```{r}
dm = da %>% group_by(species) %>% summarize(species_latitude=mean(decimalLatitude))
da = inner_join(da, dm, by="species")
```

Convert longitude to radians

```{r}
da = da %>% mutate(lonrad=pi*decimalLongitude/180)
```

Plot the locations of the occurrences.

```{r}
world = map_data("world")
plt = ggplot() + ggtitle(pclass)
plt = plt + geom_map(data=world, map=world, aes(long, lat, map_id=region, fill="grey"))
plt = plt + geom_point(data=da, aes(decimalLongitude, decimalLatitude), alpha=0.2)
plt = plt + theme(legend.position="none")
print(plt)
```

Fit a multilevel linear regression with random species intercepts. 

```{r}
m1 = lmer(decimalLatitude ~ (1 | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m1)
```
A key parameter in the output above is the standard deviation of the random day slopes, which we can extract as follows

```{r}
vc = VarCorr(m1)
print(vc)
```
We can use the multilevel model to estimate the intraclass correlation coefficient (ICC) for the species latitudes.

```{r}
sv = vc$species[1] # the species variance
rv = summary(m1)$sigma^2 # the residual variance
icc = sv / (sv + rv)
print(icc)
```
 
Below we fit a multilevel linear regression with random species intercepts and random day slopes for species.

```{r}
m2 = lmer(decimalLatitude ~ (1 + day | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m2)
```

```{r}
print(AIC(m1)-AIC(m2))
```

Below we plot the joint density of the random slope and random intercept.  The joint density is Gaussian, so we sample a large number of observations from this Gaussian distribution and plot its empirical distribution (based on 2d nonparametric density estimation).

```{r}
vc = VarCorr(m2)
sv = vc$species[1] # the species variance
va = as.matrix(vc$species)
mn = c(0, fixef(m2)[2])
vr = chol(va)
stopifnot(max(abs(va - t(vr) %*% vr)) < 1e-12)
n = 10000
z = array(rnorm(n*2), c(n, 2)) %*% vr + outer(array(1, n), mn)
z = data.frame(intercept=z[,1], slope=z[,2])
plt = ggplot(z, aes(x=intercept, y=slope)) + stat_density_2d()
print(plt)
```

Later we will conduct a formal test that the day (by species) variance is significantly positive. This requires us to fit two alternative models.

```{r}
m3 = lmer(decimalLatitude ~ (0 + day | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
summary(m3)
```

```{r}
m4 = lmer(decimalLatitude ~ (0 + day | species) + (1 | species) + day + sin(lonrad) + cos(lonrad) + sin(lonrad/2) + cos(lonrad/2), da)
```

The formal test is below, it may be very slow to run so comment it out if these results are not needed.

```{r}
#ss = exactRLRT(m3, m4, m1)
```

Estimate the marginal mean latitude at Detroit's longitude

```{r}
dp = head(da, 100)
dp = dp %>% mutate(day=seq(min(da$day), max(da$day), length.out=100))
dp = dp %>% mutate(lonrad=-pi*83/180) # Detroit longitude
dp$yy = predict(m1, dp, re.form=NA)
```

Plot the marginal mean latitude for plants observed at Detroit's longitude 

```{r}
plt = ggplot(dp, aes(x=day, y=yy)) + geom_line() + xlab("Day (x1000)") + ylab("Marginal mean latitude")
print(plt)
```

The next set of cells produces predicted "trend lines" for each species using the fitted multilevel model.  These trend lines are "partially pooled", so that for species with large sample sizes, the trend line mainly follows the observed data for that species.  But for species with small sample sizes, the intercept and slope for the species are shrunken toward the population mean intercept and slope.

First we get a predicted day 0 latitude value for each species.  Technically this is the marginal mean of the predicted day 0 latitudes, marginalizing over the longitude values (which do vary within a species but not by a lot for most species).

```{r}
dx = da
dx$day = 0
dx$Latitude0 = predict(m2, dx, re.form=NA)
dx = dx %>% group_by(species) %>% summarize(Latitude0=mean(Latitude0))
```

Plot species-level linear trends over this time range

```{r}
day1 = as.numeric(min(da$day))
day2 = as.numeric(max(da$day))
```

Extract the species random effects (random day slopes and intercepts)

```{r}
r0 = ranef(m2)$species
r0 = as.data.frame(r0)
r0$species = row.names(r0)
row.names(r0) = NULL
r0 = r0 %>% left_join(group_by(da, species) %>% summarize(species_latitude=mean(species_latitude)))
```

Use the predicted random effects (BLUPs) to predict species-level mean latitudes on the first and last day covered by the dataset.

```{r}
rr = left_join(r0, dx, by="species")
rr = rr %>% mutate(Latitude1=Latitude0+day*(day2-day1))
rr = rr %>% select(species, Latitude0, Latitude1)
```

Reshape the predictions for plotting

```{r}
pr = pivot_longer(rr, cols=Latitude0:Latitude1)
pr = rename(pr, day=name, latitude=value)
pr = pr %>% mutate(day=recode(day, Latitude0='Day0', Latitude1='Day1'))
pr = pr %>% mutate(name=ifelse(day=="Day0", day1, day2))
```

Plot species-level trends in latitude

```{r}
plt = ggplot(pr, aes(x=day, y=latitude, group=species)) + geom_line(alpha=0.2)
```

Plot the population averaged trend line in red.

```{r}
ang = -pi*83/180
f = fixef(m2)
a = f[["(Intercept)"]] + f[["sin(lonrad)"]]*sin(ang) + f[["cos(lonrad)"]]*cos(ang) +
    f["sin(lonrad/2)"]*sin(ang/2) + f["cos(lonrad/2)"]*cos(ang/2)
b = f[["day"]]
plt = plt + geom_abline(aes(intercept=a, slope=b, color="red"))
```

```{r}
plt = plt + xlab("Day (x1000)") + ylab("Species latitude")
print(plt)
```
