# ARGO data

This notebook demonstrates how we can use the concept of "data depth" to identify central and outlying points in the ARGO data.

```{r}
library(dplyr)
library(ggplot2)
```

```{r}
source("read.R")
```

The code below takes a while to run so subsample the columns.  You may need to subsample further (reduce m) if your computer is slow.

```{r}
m = 10000
ii = sample(1:dim(temp)[2], m, replace=F)
tempx = temp[,ii]
psalx = psal[,ii]
latx = lat[ii]
lonx = lon[ii]
dayx = day[ii]
```

Calculate the spatial depth of column i of x relative to the other columns of x.

```{r}
sdepth = function(i, x) {
    p = dim(x)[1]
    n = dim(x)[2]
    z = x - x[,i]
    zn = sqrt(colSums(z^2))
    zn[i] = Inf
    z = z / outer(array(1, p), zn)
    u = apply(z, 1, mean)
    return(1 - sqrt(sum(u^2)))
}
```

Calculate the L2 depth of column i of x relative to the other columns of x.

```{r}
l2depth = function(i, x) {
    p = dim(x)[1]
    n = dim(x)[2]
    z = x - x[,i]
    zn = sqrt(colSums(z^2))
    return(1e6 / (1 + mean(zn)))
}
```

Calculate the depth of every column of x relative to the other columns, using the given depth function.

```{r}
depths = function(x, dfun, progress=F) {
    n = dim(x)[2]
    d = array(0, n)
    for (i in 1:n) {
        if (progress && (i %% 200 == 0)) {
            cat(i)
            cat(".")
        }
        d[i] = dfun(i, x)
    }
    if (progress) {
        cat("\n")
    }
    return(d)
}
```

```{r}
plot_mean = function(x, pressure, ylab) {
    y = apply(x, 1, mean)
    da = data.frame(pressure=pressure, y=y)
    plt = ggplot(aes(x=pressure, y=y), data=da) + geom_line()
    plt = plt + ylab(ylab) + xlab("Pressure")
    print(plt)
}
```

Plot the mean curves for temperature and salinity.

```{r}
plot_mean(temp, pressure, "Mean temperature")
```

```{r}
plot_mean(psal, pressure, "Mean salinity")
```

Partition the depths into q equal-sized bins.  Within each bin plot the response variable (temperature or salinity) against pressure for 10 randomly selected profiles.

```{r}
depth_cut = function(x, q, pressure, ylab) {
    dp = depths(x, l2depth, progress=T)
    dq = ntile(dp, q)

    for (i in seq(1, q, 2)) {

        da = data.frame()
        for (k in c(i, i+1)) {
            ii = which(dq == k)
            jj = sample(ii, 10)
            for (j in jj) {
                dd = data.frame(pressure=pressure, y=x[,j], j=j, k=k)
                da = rbind(da, dd)
            }
        }
        da$j = as.factor(da$j)
        da$k = as.factor(da$k)

        plt = ggplot(aes(x=pressure, y=y, color=k, group=j), data=da) + geom_line()
        plt = plt + ggtitle(sprintf("Depth band %d-%d", i, i+1))
        plt = plt + ylab(ylab) + xlab("Pressure")
        print(plt)
    }
    return(dp)
}
```

Plot a small random selection of profiles from each depth interval.

```{r}
q = 20
dp_temp = depth_cut(tempx, q, pressure, "Temperature")
```

```{r}
dp_psal = depth_cut(psalx, q, pressure, "Salinity")
```

Plot the estimated conditional mean depth relative to each explanatory variable.
The bands in these plots are +/- 2 times the mean absolute deviation from the conditional mean.

```{r}
depth_correlates = function(dp, ttl) {
    dpx = data.frame(depth=dp, lat=latx, lon=lonx, day=dayx)
    for (v in c("lat", "lon", "day")) {
        xx = seq(min(dpx[[v]]), max(dpx[[v]]), length.out=100)
        m = lowess(dpx[[v]], dpx$depth)
        aresid = abs(m$y - dpx$depth)
        r = lowess(dpx[[v]], aresid)
        dh = approxfun(m$x, m$y)(xx)
        dq = approxfun(r$x, r$y)(xx)
        da = data.frame(x=xx, depth=dh, r=dq)
        f = 2
        da$ymin = dh - f*dq
        da$ymax = dh + f*dq
        plt = ggplot(aes(x=x, y=depth), data=da)
        plt = plt + geom_ribbon(aes(x=x, ymin=ymin, ymax=ymax, y=depth), fill="grey70", data=da) + geom_line()
        plt = plt + labs(x=v, y="Depth") + ggtitle(ttl)
        print(plt)
    }
}
```

```{r}
depth_correlates(dp_temp, "Temperature")
```

```{r}
depth_correlates(dp_psal, "Salinity")
```
