# Dimension reduction analysis of Argo data

The Argo data consist of paired multivariate observations for
temperature and salinity, measured simultaneously at the same physical
location at a series of depths below the ocean surface.  Each
observation consists of 100-dimensional vectors of temperature and
salinity, indexed as functions of pressure.  Here we use dimension
reduction methods including Principal Components Analysis (PCA) and
Canonical Correlation Analysis (CCA) to better understand these
profiles.

```{r}
source("read.R")

library(ggplot2)
library(dplyr)
library(dr)
```

```{r}
m = dim(temp)[1] # Number of pressure points
n = dim(temp)[2] # Number of profiles
```

Convert latitude so that the Pacific ocean doesn't wrap around the origin.

```{r}
lonx = (lon + 60) %% 360
```

Calculate the mean profiles for temperature and salinity:

```{r}
tempmean = apply(temp, 1, mean)
psalmean = apply(psal, 1, mean)
```

Plot the mean profiles:

```{r}
dp = data.frame(pressure=pressure, temp=tempmean, psal=psalmean)
plt = ggplot(dp, aes(x=pressure, y=temp)) + geom_line()
print(plt)

plt = ggplot(dp, aes(x=pressure, y=psal)) + geom_line()
print(plt)
```

Center the profiles:

```{r}
tempc = temp - outer(tempmean, array(1, n))
psalc = psal - outer(psalmean, array(1, n))
```

Get the principal components

```{r}
cc = cov(t(tempc))
temp_ee = eigen(cc, symmetric=T)
temp_eigval = temp_ee$values
temp_eigvec = temp_ee$vectors

cc = cov(t(psalc))
psal_ee = eigen(cc, symmetric=T)
psal_eigval = psal_ee$values
psal_eigvec = psal_ee$vectors
```

As in any factor analysis, the loadings are only identified up tgo
sign.  To ease interpretation, the function below flips the PC loading
vectors if needed so that the majority of loadings are positive.

```{r}
flip = function(eigvec) {
  for (j in 1:dim(eigvec)[2]) {
      if (sum(eigvec[,j] < 0) > sum(eigvec[,j] > 0)) {
        eigvec[,j] = -eigvec[,j]
      }
  }
  return(eigvec)
}

temp_eigvec = flip(temp_eigvec)
psal_eigvec = flip(psal_eigvec)
```

Scores for the dominant PC's

```{r}
tscores = t(tempc) %*% temp_eigvec[,1:10]
pscores = t(psalc) %*% psal_eigvec[,1:10]
```

Plot the loadings and means plus/minus multiples of the loadings.

```{r}
plot_loadings = function(ymean, scores, eigvec, ti) {

  da = data.frame(pressure=pressure, ymean=ymean)

  # Make several plots depicting the loading patterns.
  for (j in 1:5) {
      # Plot the loadings
      da = data.frame(pressure=pressure, loading=eigvec[,j])
      plt = ggplot(aes(x=pressure, y=loading), data=da) + geom_line()
      plt = plt + labs(x="Pressure", y=sprintf("%s loading", ti))
      plt = plt + ggtitle(sprintf("PC %d", j))
      print(plt)

      # Plot the mean profile +/- multiples of the loadings
      da = data.frame()
      s = sd(scores[,j])
      for (f in c(-1, 0, 1)) {
        dx = data.frame(pressure=pressure, profile=ymean+f*s*eigvec[,j], f=f)
        da = rbind(da, dx)
      }
      da$f = as.factor(da$f)
      plt = ggplot(aes(x=pressure, y=profile, color=f, group=f), data=da) + geom_line()
      plt = plt + labs(x="Pressure", y=ti)
      plt = plt + ggtitle(sprintf("PC %d", j))
      print(plt)
  }
}
```

```{r}
plot_loadings(tempmean, tscores, temp_eigvec, "Temperature")
```

```{r}
plot_loadings(psalmean, pscores, psal_eigvec, "Salinity")
```

Plot the j^th PC score against the k^th feature.

```{r}
plot_pcscores = function(j, k) {

    xx = seq(min(Y[,k]), max(Y[,k]), length=100)
    m = lowess(Y[,k], tscores[,j])
    mf = approxfun(m$x, m$y)
    f = abs(m$y - tscores[,j])
    r = lowess(Y[,k], f)
    rf = approxfun(r$x, r$y)
    da = data.frame(x=xx, y=mf(xx), r=rf(xx))

    f = 2
    da$y1 = da$y - f*da$r
    da$y2 = da$y + f*da$r
    plt = ggplot(aes(x=x, y=y), data=da)
    plt = plt + labs(x=fn[k], y=sprintf("PC %d score", j))
    plt = plt + geom_ribbon(aes(x=x, ymin=y1, ymax=y2), fill="grey70")
    plt = plt + geom_line()
    print(plt)
}
```

```{r}
fn = c("Latitude", "Longitude", "Day")
Y = cbind(lat, lon, day)
```

```{r}
for (j in 1:3) {
    for (k in 1:3) {
        plot_pcscores(j, k)
    }
}
```

Analogous to 'flip_pca' above, this function flip the signs in two
sets of CCA loadings so that the majority of loadings are positive.
This may make them easier to interpret in a plot.

```{r}
flip_cca = function(xcoef, ycoef) {
    for (j in 1:dim(xcoef)[2]) {
        if (mean(xcoef[,j] > 0) + mean(ycoef[,j] > 0) < 1) {
            xcoef[,j] = -xcoef[,j]
            ycoef[,j] = -ycoef[,j]
        }
    }
    return(list(xcoef=xcoef, ycoef=ycoef))
}
```

We could apply CCA directly to the 100-dimensional data, but even with
our large sample size this would result in overfitting.  Therefore, we
use a combined PCA/CCA approach to relate temperature and
salinity. First, PCA is used to reduce the dimension and constrain the
analysis to the subspace capturing the greatest fraction of variation.
Then, CCA is used in this subspace to identify the maximally
correlated components.  The function below produces plots for the
results of this approach, for a sequence of dimensions 'q' defining
the extent of PCA dimension reduction.

```{r}
plot_cca = function() {
    X = t(tempc)
    Y = t(psalc)
    svx = svd(X)
    svy = svd(Y)

    for (q in c(1, 2, 5, 10)) {
        cc = cancor(svx$u[,1:q], svy$u[,1:q])
        rr = cc$cor
        print("Canonical correlations: ")
        print(rr)

        # Passing nrow and ncol is needed to avoid
        # incorrect behavior when q=1.
        ddx = diag(svx$d[1:q], q, q)
        ddy = diag(svy$d[1:q], q, q)

        # Map the loadings back to the original coordinates
        xcoef = svx$v[,1:q] %*% solve(ddx, cc$xcoef)
        ycoef = svy$v[,1:q] %*% solve(ddy, cc$ycoef)
        ll = flip_cca(xcoef, ycoef)
        xcoef = ll$xcoef
        ycoef = ll$ycoef

        da = data.frame(temp_weights=xcoef[,1], psal_weights=ycoef[,1], pressure=pressure)
        plt = ggplot(aes(x=pressure, y=temp_weights), data=da) + geom_line()
        plt = plt + labs(x="Pressure", y="Temperature coefficient")
        plt = plt + ggtitle(sprintf("CCA with q=%d components, r=%.2f", q, rr[1]))
        if (min(ycoef[,1]) > 0) {
            plt = plt + ylim(0, NA)
        }
        print(plt)
        plt = ggplot(aes(x=pressure, y=psal_weights), data=da) + geom_line()
        plt = plt + labs(x="Pressure", y="Salinity coefficient")
        plt = plt + ggtitle(sprintf("CCA with q=%d components, r=%.2f", q, rr[1]))
        if (min(ycoef[,1]) > 0) {
            plt = plt + ylim(0, NA)
        }
        print(plt)
    }
}
```

```{r}
plot_cca()
```

```{r}
plot_cc_map = function(q=2) {
    X = t(tempc)
    Y = t(psalc)
    svx = svd(X)
    svy = svd(Y)
    cc = cancor(svx$u[,1:q], svy$u[,1:q])

    # Passing nrow and ncol is needed to avoid
    # incorrect behavior when q=1.
    ddx = diag(svx$d[1:q], q, q)
    ddy = diag(svy$d[1:q], q, q)

    xx = svx$u[,1:q] %*% cc$xcoef
    yy = svy$u[,1:q] %*% cc$ycoef

    grp = array(0, dim(xx)[1])
    j1 = xx[,1] >= quantile(xx[,1], 0.9)
    j2 = yy[,1] >= quantile(yy[,1], 0.9)
    grp[j1 & j2] = 1

    k1 = xx[,1] <= quantile(xx[,1], 0.1)
    k2 = yy[,1] <= quantile(yy[,1], 0.1)
    grp[k1 & k2] = 2

    dp = data.frame(lat=lat, lon=lonx, grp=grp)
    dp = dp %>% filter(grp > 0)
    dp$grp = as.factor(dp$grp)
    plt = ggplot(dp, aes(x=lon, y=lat, by=grp, color=grp)) + geom_point(alpha=0.2)
    print(plt)

    # TODO plot loadings
}

plot_cc_map()
```

# Dimension reduction regression

SIR is a dimension reduction regression (DR) method.  Here we will use
it to identify factors within the temperature data that predict
latitude.

```{r}
plot_sir = function() {

    X = t(tempc)

    # Due to high-dimensionality, we project the temperature data to
    # a limited number of PC's before using SIR to estimate the coefficients.
    q = 3
    svx = svd(X)
    dd = dr.compute(svx$u[,1:q], lat, array(1, length(lat)))
    ddx = diag(svx$d[1:q], q, q)
    b = svx$v[,1:q] %*% solve(ddx, dd$evectors)

    # Plot the loadings
    da = data.frame(pressure=c(pressure, pressure, pressure))
    da$dir = c(b[,1], b[,2], b[,3])
    oo = array(1, length(pressure))
    da$group = c(1*oo, 2*oo, 3*oo)
    da$group = as.factor(da$group)
    plt = ggplot(aes(x=pressure, y=dir, by=group, color=group), data=da) + geom_line()
    plt = plt + labs(x="Pressure", y="Temperature loading")
    plt = plt + ggtitle(sprintf("SIR with q=%d PC components", q))
    print(plt)

    # Plot the scores against latitude, longitude, and day.
    scores = X %*% b[,1:3]
    for (v in c("lat", "lon", "day")) {
        for (j in 1:3) {
            da = data.frame(lat=lat, lon=lon, day=day, s=scores[,j])
            plt = ggplot(aes(x=!!sym(v), y=s), data=da) + geom_point(alpha=0.1)
            plt = plt + labs(x=v, y=sprintf("SIR component %d", j))
            print(plt)
        }
    }
}
```

```{r}
da = plot_sir()
```
